Viscous Transport in Eroding Porous Media

### 2019-10-03 docs/transport/erosion_transport.tex
small changes to intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998} and velocity scales
can be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}.
Moreover, for a single porous geometry, the pore sizes and velocities
can range over several orders of magnitude.  Numerical methods that
resolve this range of scales offer the ability to: (i) characterize
dispersion~\cite{saf1959}, (ii) quantify
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and
(iii) potentially develop meaningful constitutive relationships that
link the microscopic and macroscopic
realms~\cite{mil-chr-imh-mcb-ped1998}.  Examples of coarse-grained
models for porous media flow include pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is further complicated when boundaries dynamically
evolve in response to the fluid flow. This type of coupling between the
geometry and the flow occurs, for example, in applications involving
melting~\cite{bec-vis1998, rycroft2016asymmetric, jambon2018singular,
favier2019rayleigh, morrow2019moving},
dissolution~\cite{kan-zha-che-he2002, mac2015shape, moo2017,
wykes2018self}, deposition~\cite{joh-eli1995, hewett2018modelling},
biofilm growth~\cite{tan-val-wer2015}, and crack
formation~\cite{cho2019crack}. We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, and
industrial applications~\cite{berhanu2012shape, hewett2017evolution,
lachaussee2018competitive, lopez2018cfd, allen2019sde, amin2019role}.
When a porous geometry erodes, certain qualitative characteristics are
unveiled that affect transport through the geometry.  For example, an
eroded geometry may contain channels of high porosity, which, though few
in number and modest in volume fraction, transmit a large portion of the
flux \cite{qua-moo2018}. This arrangement results in velocities that vary over several
orders of magnitude~\cite{all-hea-lab-rei2002}.  Moreover,
channelization creates heterogeneous and anisotropic medium properties,
which affect the transport of tracers such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterization of
tracer transport through the resulting eroded geometries. Our modeling
efforts build on previous work~\cite{ris-moo-chi-she-zha2012,
moore2013self, moo2017}, in particular recent numerical methods
developed to simulate erosion in the Stokes-flow
regime~\cite{qua-moo2018}. We, however, make key improvements to the
numerical methods from~\cite{qua-moo2018} to enable simulations of more
realistic, dense suspensions of eroding grains
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To
characterize transport through these configurations, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration method is unchanged from
previous work~\cite{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be triggered by changes in
sign of the shear stress, and we use a stable second-order Runge-Kutta
method applied to the {\thL} coordinates~\cite{hou-low-she1994} of the
eroding grains.  In this work, we introduce a new quadrature method to
resolve dense suspensions.  The accuracy of the
trapezoid rule, which was used in previous work, is adequate for
bodies that are sufficiently separated~\cite{tre-wei2014}.  However, it
loses accuracy whenever grains are in close contact.  One of the
earliest works to address quadrature methods for nearly-singular
integrands was developed by Baker and Shelley~\cite{bak-she1986}, and in
recent years, many other near-singular integration schemes have
followed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\cite{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate shear stress and
fluid vorticity.

@@

Once the streamlines are computed, we characterize transport by
analyzing three different metrics: the tortuosity, the anomalous
dispersion, and the pore size distribution.  The local tortuosity of a
streamline that connects the inlet to the outlet is defined as the
streamline's length normalized by the linear inlet-to-outlet distance.
In porous media, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008} or even
2~\cite{dud-koz-mat2011}, depending on several factors such as the
porosity. The tortuosity of a geometry is defined by averaging the local
tortuosity over all streamlines initialized at the inlet, and the
geometry's tortuosity characterizes average particle
motions~\cite{hak-com-den2019}.  To characterize spreading, the fluid
dispersion is computed by calculating the variance of the streamline
lengths. In porous media, this spreading is often
super-dispersive~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. Since anomalous dispersion results from
streamlines spending time in both the high and low velocity
regimes~\cite{ber-sch2001}, it is crucial to accurately resolve
streamlines near grain boundaries, as achieved in this work.  Finally,
the distribution of pore sizes is required to quantify velocity
distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019},
connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}. Hence, we
construct the pore-size distribution throughout the erosion process.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the erosion model that is
described in more detail in previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry, transport, and dispersion. Section~\ref{sec:method} describes
the numerical methods, with special attention paid to the new quadrature
method for computing the shear stress and the vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
Section~\ref{sec:conclusions}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998} and velocity scales
can be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}.
Moreover, for a single porous geometry, the pore sizes and velocities
can range over several orders of magnitude.  Numerical methods that
resolve this range of scales offer the ability to: (i) characterize
dispersion~\cite{saf1959}, (ii) quantify
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and
(iii) develop meaningful constitutive relationships that link the
microscopic and macroscopic realms~\cite{mil-chr-imh-mcb-ped1998}.
Examples of coarse-grained models for porous media flow include
pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, pore and flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is further complicated when boundaries dynamically
evolve in response to the fluid flow. This coupling between the geometry
and the flow occurs, for example, in applications involving
melting~\cite{bec-vis1998, rycroft2016asymmetric, jambon2018singular,
favier2019rayleigh, morrow2019moving},
dissolution~\cite{kan-zha-che-he2002, mac2015shape, moo2017,
wykes2018self}, deposition~\cite{joh-eli1995, hewett2018modelling},
biofilm growth~\cite{tan-val-wer2015}, and crack
formation~\cite{cho2019crack}. We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, and
industrial applications~\cite{berhanu2012shape, hewett2017evolution,
lachaussee2018competitive, lopez2018cfd, allen2019sde, amin2019role}.
When a porous geometry erodes, certain qualitative characteristics are
unveiled that affect transport through the geometry.  For example, an
eroded geometry may contain channels of high porosity, which, though few
in number and modest in volume fraction, transmit a large portion of the
flux \cite{qua-moo2018}. This arrangement results in velocities that
vary over several orders of magnitude~\cite{all-hea-lab-rei2002}.
Moreover, channelization creates heterogeneous and anisotropic medium
properties, which affect the transport of tracers such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterization of
tracer transport through the resulting eroded geometries. Our modeling
efforts build on previous work~\cite{ris-moo-chi-she-zha2012,
moore2013self, moo2017}, in particular recent numerical methods
developed to simulate erosion in the Stokes-flow
regime~\cite{qua-moo2018}. We, however, make key improvements to the
numerical methods from~\cite{qua-moo2018} to enable simulations of more
realistic, dense suspensions of eroding grains
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  Then, to
characterize transport through these configurations, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration method is unchanged from
previous work~\cite{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be triggered by changes in
sign of the shear stress, and we use a stable second-order Runge-Kutta
method applied to the {\thL} coordinates~\cite{hou-low-she1994} of the
eroding grains.  In this work, we introduce a new quadrature method to
resolve dense suspensions.  The accuracy of the
trapezoid rule, which was used in previous work, is adequate for
bodies that are sufficiently separated~\cite{tre-wei2014}.  However, it
loses accuracy whenever grains are in close contact.  One of the
earliest works to address quadrature methods for nearly-singular
integrands was developed by Baker and Shelley~\cite{bak-she1986}, and in
recent years, many other near-singular integration schemes have
followed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\cite{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate the shear stress and
the fluid vorticity.

@@

Once the streamlines are computed, we characterize transport by
analyzing three different metrics: the tortuosity, the anomalous
dispersion, and the pore size distribution.  The local tortuosity of a
streamline that connects the inlet to the outlet is defined as the
streamline's length normalized by the linear inlet-to-outlet distance.
In porous media, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008} or even
2~\cite{dud-koz-mat2011}, depending on several factors such as the
porosity. The tortuosity of a geometry is defined by averaging the local
tortuosity over all streamlines initialized at the inlet, and the
geometry's tortuosity characterizes average particle
motions~\cite{hak-com-den2019}.  To characterize spreading, the fluid
dispersion is computed by calculating the variance of the streamline
lengths. In porous media, this spreading is often
super-dispersive~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. Since anomalous dispersion results from
streamlines spending time in both the high and low velocity
regimes~\cite{ber-sch2001}, it is crucial to accurately resolve
streamlines near grain boundaries, as achieved in this work.  Finally,
we construct the pore-size distribution throughout the erosion process.
These distributions are required to quantify velocity
distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019},
connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the erosion model that is
described in more detail in previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry and transport. Section~\ref{sec:method} describes the numerical
methods, with special attention paid to the new quadrature method for
computing the shear stress and the vorticity.  Section~\ref{sec:results}
presents numerical examples for a variety of dense packings of bodies.
Finally, concluding remarks are made in Section~\ref{sec:conclusions}.
